%%
% 摘要
%%

\cabstract{

随着位置服务与移动互联网的发展，下一个兴趣点（Point-of-Interest, POI）预测已成为智能出行与个性化推荐的重要问题。现有方法在该任务上面临三方面不足：其一，传统序列与图模型通常采用统一转移机制，难以刻画同一POI在不同时段的行为差异，且仅建模"当前点去哪里"而忽略"目标点从哪里来"的方向性信息，导致路径预测存在偏置；其二，大语言模型（LLM）将经纬度作为文本token处理，地理上相邻的坐标在离散token空间中可能被映射为差异较大的符号序列，致使空间连续性被破坏，容易产生地理不一致的预测；其三，POI之间的高阶转移关系蕴含在图结构中，仅依赖文本上下文难以将这类结构先验有效注入LLM语义空间，制约了模型对未显式出现候选的推断能力。针对上述问题，本文围绕"大小模型协同学习"展开研究，构建融合时空结构先验与语义推理能力的统一框架。

方法上，针对时间异质性与转移方向性不足，本文在小模型侧构建时间增强序列动态图与双向转移建模，按时段划分转移子图并同时学习"转出-转入"关系；针对LLM空间连续性缺失，设计地理坐标注入模块（GCIM），通过层级离散编码与连续频域编码的融合将地理约束以结构化方式注入语义空间；针对转移先验难以注入LLM，设计POI对齐模块（PAM），建立从图结构表征到LLM语义空间的可学习映射通道，使高阶转移关系在语义侧可被读取和利用；在训练策略上采用"先对齐、后协同"的两阶段流程，以参数高效微调控制开销。在公开签到数据及其城市子集上的实验表明，所提方法在Acc@1、MRR等核心指标上均优于现有序列、图与大语言模型基线；其中，TSPM在Gowalla上的Acc@1相对最强非本文基线提升5.49\%，GA-LLM在NYC上的Acc@1提升18.27\%。消融实验与诊断分析进一步验证了各模块的独立贡献，并显示该框架在精度提升与工程可部署性之间取得了稳定平衡。
}

\ckeywords{下一个兴趣点推荐；大小模型协同；时空建模；大语言模型；参数高效微调}

\eabstract{

Next point-of-interest (POI) prediction is a key task for intelligent mobility and personalized recommendation. Existing methods suffer from three limitations. First, conventional sequential and graph-based models adopt a unified transition mechanism that fails to capture the behavioral differences of the same POI across time periods, and model only ``where the current point leads to'' while ignoring ``where the target point is typically reached from,'' introducing directional bias in path prediction. Second, large language models (LLMs) tokenize geographic coordinates as plain text, so that spatially adjacent locations may be mapped to dissimilar token sequences, breaking spatial continuity and causing geographically inconsistent predictions. Third, high-order transition patterns are embedded in graph structures and cannot be effectively injected into the LLM semantic space through text context alone, limiting the model's ability to infer candidates that do not explicitly appear in the input. To address these issues, this thesis develops a unified small-large collaborative learning framework that integrates spatio-temporal structural priors with semantic reasoning.

On the methodology side, to address temporal heterogeneity and insufficient transition directionality, a time-enhanced sequence-based dynamic graph with bidirectional transfer modeling is introduced on the small-model branch, which partitions transition sub-graphs by time slot and jointly learns outgoing and incoming preferences. To tackle the lack of spatial continuity in LLMs, a Geographic Coordinate Injection Module (GCIM) is designed to fuse hierarchical discrete encoding with continuous Fourier encoding and inject geographic constraints into the semantic space in a structured manner. To bridge the gap in transfer-prior injection, a POI Alignment Module (PAM) establishes a learnable mapping from graph-based structural representations to the LLM semantic space, enabling high-order transition knowledge to be accessed and utilized on the semantic side. A two-stage training strategy (``alignment first, collaboration second'') with parameter-efficient fine-tuning is adopted to control computational overhead. Experiments on public check-in datasets and their city-level subsets show that the proposed method consistently outperforms sequential, graph-based, and LLM baselines on Acc@1, MRR, and related metrics. In particular, TSPM improves Acc@1 by 5.49\% on Gowalla, and GA-LLM improves Acc@1 by 18.27\% on NYC over the strongest non-ours baselines. Ablation and diagnostic studies further verify the independent contribution of each module and confirm a stable trade-off between predictive accuracy and deployment cost.
}

\ekeywords{Next POI Recommendation; Collaborative Learning; Spatio-temporal Modeling; Large Language Models; Parameter-efficient Fine-tuning}
